Using HbA1c and Fasting BG to Predict Development of CFRD

### 2025-03-09 Christine Chan/CF Registry/cf_registry_CFRD_prediction.qmd
nonconformable errors

```{r setup}
#| include: false
library(tidyverse)
library(arsenal)
library(childsds)
library(survival)
library(ggsurvfit)
library(ggstats)
library(gtsummary)
knitr::opts_knit$set(root.dir = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry")
```

@@

@@

# CFRD by OGTT two hour
annual$cfrd_ogtt <- factor(annual$A_twohour_bloodglucose >= 200,
  levels = c(F, T), labels = c("CFRD-", "CFRD+")
# CFRD by registry definition
annual$cfrd_registry <- factor(annual$cfrd_status_annual,
  levels = c(1:5), labels = c("CFRD-", "CFRD-", "CFRD+", "CFRD-", "CFRD-")

@@

annual$cfrd_registry[is.na(annual$cfrd_registry)] <- "CFRD-"
annual$cfrd_ogtt[is.na(annual$cfrd_ogtt)] <- "CFRD-"

@@

# Convert BMI to percentile (assume age > 20 is 20)
annual$bmi_perc <- sds(annual$A_bmivalue,
  age = ifelse(annual$Age_YrEnd > 20, 20, annual$Age_YrEnd),
  sex = annual$Gender, item = "bmi", type = "perc", male = "M", female = "F",
  ref = cdc.ref
) * 100
# Select columns

@@

    eDWID, ReviewYear, Race, Hispanicrace, Gender, F508, MutClass,
    A_Vx770, A_VX809comb, Age_YrEnd, A_bmivalue, A_bmipercentile, bmi_perc,
    A_FEV1_pct_predicted, NumPulmExacerbation, A_IsOnEnzymes,
    A_fast_bloodglucose, A_hgba1c, cfrd_registry, cfrd_status_annual, cfrd_ogtt,
    A_twohour_bloodglucose

@@

    Age_YrEnd = round(Age_YrEnd, 2),
    cfrd_registry_age = first(na.omit(Age_YrEnd[cfrd_registry == "CFRD+"])),
    cfrd_ogtt_age = first(na.omit(Age_YrEnd[cfrd_ogtt == "CFRD+"]))

@@

annual$cfrd_registry[annual$Age_YrEnd < annual$cfrd_registry_age] <- "CFRD-"
annual$cfrd_registry[annual$Age_YrEnd >= annual$cfrd_registry_age] <- "CFRD+"
annual$cfrd_ogtt[annual$Age_YrEnd < annual$cfrd_ogtt_age] <- "CFRD-"
annual$cfrd_ogtt[annual$Age_YrEnd >= annual$cfrd_ogtt_age] <- "CFRD+"
# Split into one dataset for registry and one for OGTT
registry <- annual %>%
  filter(Age_YrEnd <= cfrd_registry_age) %>%
  select(eDWID:A_hgba1c, cfrd_registry)
  select(eDWID:A_hgba1c, cfrd_ogtt, cfrd_ogtt_age)
# Create start times
registry$Age_YrStart <- registry$Age_YrEnd - 1
ogtt$Age_YrStart <- ogtt$Age_YrEnd - 1

@@

@@

  Age_YrStart = "Age at Year Start", Age_YrEnd = "Age at Year End",

@@

  NumPulmExacerbation = "Admissions for PEx"

@@

# Basic formula
xvars <- c(
  "Gender", "Race", "Hispanicrace", "MutClass", "A_IsOnEnzymes",
  "A_FEV1_pct_predicted", "NumPulmExacerbation", "A_corticosteroids1"

@@

f <- ~ A_hgba1c + A_fast_bloodglucose + Gender + Race + Hispanicrace + MutClass +
  A_IsOnEnzymes + A_FEV1_pct_predicted + NumPulmExacerbation

@@

# By registry CFRD status

## Participant demographics

Categorical variables are from either the first year with CFRD or the most recent year for those without CFRD.

```{r results='asis'}
v1 <- registry %>%
  group_by(eDWID) %>%
  slice_max(ReviewYear)
t1 <- tableby(cfrd_registry ~ Race + Hispanicrace + Gender + F508 + MutClass +
  Age_YrEnd + A_bmipercentile + A_hgba1c, data = v1)
summary(t1, pfootnote = T)
```

## Kaplan-Meier plots

```{r}
# Plot
survfit2(Surv(Age_YrStart, Age_YrEnd, as.numeric(cfrd_registry)) ~ 1,
  data = registry
) %>%
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall CFRD- probability"
  ) +
  add_confidence_interval() +
  add_risktable()
# Median "survival" time
survfit2(Surv(Age_YrStart, Age_YrEnd, as.numeric(cfrd_registry)) ~ 1,
  data = registry
) %>%
  tbl_survfit(
    probs = 0.5,
    label_header = "**Median time to CFRD+ (95% CI)**"
  )
```

## Cox PH model

```{r}
cph <- coxph(
  update(f, Surv(time = Age_YrStart, time2 = Age_YrEnd, event = as.numeric(cfrd_registry)==2) ~ .),
  data = registry, cluster = eDWID
)
cph %>% tbl_regression(exponentiate = T)
plot(cox.zph(cph), resid = F)
cox.zph(cph)
ggcoef_table(cph,
  exponentiate = T, show_p_values = F, add_reference_rows = T, signif_stars = F,
  colour = F
)

d = model.frame(cph)
d$pred <- predict(cph, type = "survival")
```

# By CFRD status based on 2 hour OGTT >= 200

## Participant demographics

Categorical variables are from either the first year with CFRD or the most recent year for those without CFRD.

```{r results='asis'}
v1 <- ogtt %>%
  group_by(eDWID) %>%
  slice_max(ReviewYear)
t1 <- tableby(cfrd_ogtt ~ Race + Hispanicrace + Gender + F508 + MutClass +
  Age_YrEnd + A_bmipercentile + A_hgba1c, data = v1)
summary(t1, pfootnote = T)
```

## Kaplan-Meier plots

```{r}
# Plot
survfit2(Surv(Age_YrStart, Age_YrEnd, as.numeric(cfrd_ogtt)) ~ 1,
  data = ogtt
) %>%
  ggsurvfit() +
  labs(
    x = "Age",
    y = "Overall CFRD- probability"
  ) +
  add_confidence_interval() +
  add_risktable()
# Median "survival" time
survfit2(Surv(Age_YrStart, Age_YrEnd, as.numeric(cfrd_ogtt)) ~ 1,
  data = ogtt
) %>%
  tbl_survfit(
    probs = 0.5,
    label_header = "**Median time to CFRD+ (95% CI)**"
  )
```

## Cox PH model

```{r}
cph <- coxph(
  update(f, Surv(time = Age_YrStart, time2 = Age_YrEnd, event = as.numeric(cfrd_ogtt)) ~ .),
  data = ogtt, cluster = eDWID
)
cph %>% tbl_regression(exponentiate = T)
plot(cox.zph(cph), resid = F)
cox.zph(cph)
ggcoef_table(cph,
  exponentiate = T, show_p_values = F, add_reference_rows = T, signif_stars = F,
  colour = F
)
```

>>>>

```{r setup}
#| include: false
library(survival)
library(nlme)
library(JM)
library(dplyr)
library(tidyr)
knitr::opts_knit$set(root.dir = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/CF Registry")
```

@@

# Exclude T1D and T2D
annual <- annual %>% filter(!cfrd_status_annual %in% c(4, 5))
annual$cfrd_registry <- as.numeric(annual$cfrd_status_annual == 3)
annual$cfrd_ogtt <- as.numeric(annual$A_twohour_bloodglucose >= 200)

@@

@@

annual$cfrd_registry[is.na(annual$cfrd_registry)] <- 0
annual$cfrd_ogtt[is.na(annual$cfrd_ogtt)] <- 0

@@

# Get age at first CFRD (by OGTT and registry)

@@

@@

    cfrd_registry_age = first(na.omit(Age_YrEnd[cfrd_registry == 1])),
    cfrd_ogtt_age = first(na.omit(Age_YrEnd[cfrd_ogtt == 1]))

@@

annual$cfrd_registry[annual$Age_YrEnd < annual$cfrd_registry_age] <- 0
annual$cfrd_registry[annual$Age_YrEnd >= annual$cfrd_registry_age] <- 1
annual$cfrd_ogtt[annual$Age_YrEnd < annual$cfrd_ogtt_age] <- 0
annual$cfrd_ogtt[annual$Age_YrEnd >= annual$cfrd_ogtt_age] <- 1

@@

annual$A_IsOnEnzymes <- factor(annual$A_IsOnEnzymes,
  levels = 1:0, labels = c("Yes", "No")
# Calculate bmi value
annual$bmi <- annual$A_weight / ((annual$A_height / 100)^2)
annual$bmi <- coalesce(annual$A_bmivalue, annual$bmi)

@@

  Age_YrEnd = "Age at Year End",

@@

  NumPulmExacerbation = "Admissions for PEx",
  A_IsOnEnzymes = "Pancreatic enzyme use", bmi = "BMI value"

@@

# List of predictor variables
predictors <- c(
  "Gender", "Race", "Hispanicrace", "A_IsOnEnzymes",
  "NumPulmExacerbation", "A_FEV1_pct_predicted", "bmi"

@@

# Make survival datasets for registry and OGTT diagnoses
    eDWID, ReviewYear, Age_YrEnd, A_hgba1c, all_of(predictors), cfrd_ogtt
  drop_na() %>%
  distinct()
surv_ogtt <- annual %>%
  slice_max(row_number()) %>%
  select(eDWID, ReviewYear, Age_YrEnd, all_of(predictors), cfrd_ogtt) %>%
  drop_na() %>%
  distinct()
ogtt_include <- intersect(ogtt$eDWID, surv_ogtt$eDWID)

@@

# Joint models

## CFRD diagnosis by OGTT

```{r}
# Formulas
cox_f <- as.formula(paste0(
  "Surv(time = Age_YrEnd, event = cfrd_ogtt)~",
  paste0(predictors, collapse = "+")
))
lme_f <- as.formula(paste0(
  "A_hgba1c~Age_YrEnd+",
  paste0(predictors, collapse = "+")
))
# Fit Cox model first, and then pull the clusters
cox_fit <- coxph(cox_f,
  cluster = eDWID, x = T, model = T,
  data = surv_ogtt[surv_ogtt$eDWID %in% ogtt_include, ]
)
lme_fit_a1c <- lme(lme_f,
  random = ~ Age_YrEnd | eDWID, na.action = na.omit,
  data = ogtt[ogtt$eDWID %in% ogtt_include, ]
)
# Combine
joint_fit <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd")
```
